---
title: "Latent class growth curves for Stress"
author: "Setareh"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyr)
library(dplyr)
library(ggplot2)
library(lcmm)
```

```{r}
dta_raw <- read.csv("raw40 subjects_HRV all conditions.csv", sep = ";")
```


# Visualization of individual growth curve

```{r}
plot_data <- dta_raw %>%  dplyr::select("subject", "Stress_rest1", "Stress_iat","Stress_conf", "Stress_masc" ,"Stress_rest2") %>% pivot_longer(.,names_to="time", names_prefix="Stress_",cols=starts_with("Stress_"), values_to="Stress")


ggplot(plot_data, aes(x=factor(plot_data$time, ordered=T,levels=c("rest1","iat", "conf","masc","rest2")), y=Stress, col=as.factor(subject)))+ geom_path(group=1) +geom_point()+theme(legend.position = "right") + xlab("Measurement time")
```

\newpage

# Model comparison


```{r}
data_long  <- dta_raw %>%  dplyr::select("subject", "BMI" , "age", "Stress_rest1", "Stress_iat","Stress_conf", "Stress_masc" ,"Stress_rest2") %>% pivot_longer(.,names_to="time", names_prefix="Stress_",cols=starts_with("Stress_"), values_to="Stress")

# time is kept categorical (5 phases)
data_long$time[data_long$time=="rest1"] <- 1
data_long$time[data_long$time=="iat"] <- 2
data_long$time[data_long$time=="conf"] <- 3
data_long$time[data_long$time=="masc"] <- 4
data_long$time[data_long$time=="rest2"] <- 5

```


```{r GC}
set.seed(522024)
lcga1 <- hlme(Stress ~ time , subject = "subject", ng=1, data=data_long )

lcga2 <- gridsearch(rep=200,maxiter = 10,minit = lcga1, m=hlme(Stress ~ time, subject = "subject", ng=2, data=data_long, mixture = ~ time ))

lcga3 <- gridsearch(rep=200,maxiter = 10,minit = lcga1, m=hlme(Stress ~ time, subject = "subject", ng=3, data=data_long, mixture = ~ time ))

lcga4 <- gridsearch(rep=200,maxiter = 10,minit = lcga1, m=hlme(Stress ~ time, subject = "subject", ng=4, data=data_long, mixture = ~ time ))

lcga5 <- gridsearch(rep=200,maxiter = 10,minit = lcga1, m=hlme(Stress ~ time, subject = "subject", ng=5, data=data_long, mixture = ~ time ))
```



```{r RI}
set.seed(522024)
gmm1 <- hlme(Stress ~ time , subject = "subject", random = ~ 1, ng=1, data=data_long )

gmm2 <- gridsearch(rep=200,maxiter = 10,minit = gmm1, m=hlme(Stress ~ time, subject = "subject", random = ~ 1, ng=2, data=data_long, mixture = ~ time ,nwg=T))

gmm3 <- gridsearch(rep=200,maxiter = 10,minit = gmm1, m=hlme(Stress ~ time, subject = "subject", random = ~ 1, ng=3, data=data_long, mixture = ~ time ,nwg=T ))

gmm4 <- gridsearch(rep=200,maxiter = 10,minit = gmm1, m=hlme(Stress ~ time, subject = "subject", random = ~ 1, ng=4, data=data_long, mixture = ~ time ,nwg=T))
```



```{r RS}
set.seed(522024)
gmm1_2 <- hlme(Stress ~ time , subject = "subject", random = ~ 1 + time, ng=1, data=data_long )

gmm2_2 <- gridsearch(rep=200,maxiter = 10,minit = gmm1_2, m=hlme(Stress ~ time, subject = "subject", random = ~ 1 + time, ng=2, data=data_long, mixture = ~ time ,nwg=T))

gmm3_2 <- gridsearch(rep=200,maxiter = 10,minit = gmm1_2, m=hlme(Stress ~ time, subject = "subject", random = ~ 1 + time, ng=3, data=data_long, mixture = ~ time ,nwg=T))

gmm4_2 <- gridsearch(rep=200,maxiter = 10,minit = gmm1_2, m=hlme(Stress ~ time, subject = "subject", random = ~ 1 + time, ng=4, data=data_long, mixture = ~ time ,nwg=T))
```

```{r}
pred <- predictClass(gmm2_2,newdata = data_long)
save(lcga1, lcga2, lcga3, lcga4, lcga5, gmm1, gmm2, gmm3, gmm4, gmm1_2, gmm2_2, gmm3_2, gmm4_2, pred, file="lcmm_Stress.RData")
```


\newpage

```{r}
cls <- pred[, c("subject", "class")]
names(cls)[2] <- "class_Stress"
dta_raw <- merge(dta_raw, cls, by = "subject")
```


## LCGA for Stress

Latent Class Growth Analysis, 1 to 5 classes.


```{r}
summarytable(lcga1,lcga2,lcga3, lcga4, lcga5)
```

## GMM(RI) for Stress

Growth Mixture Modelling with class-specific random intercept, 1 to 4 classes.


```{r}
summarytable(gmm1,gmm2,gmm3, gmm4)
```

## GMM(RS) for Stress

Growth Mixture Modelling with class-specific random intercept and random time effects, 1 to 4 classes.


```{r}
summarytable(gmm1_2,gmm2_2,gmm3_2, gmm4_2)
```



\newpage

- The best model is the 2-class GMM with random intercept and random time effects.

```{r}
summary(gmm2_2)
```

\newpage

```{r}
plot_data <- dta_raw %>%  dplyr::select("subject", "class_Stress" ,"Stress_rest1", "Stress_iat","Stress_conf", "Stress_masc" ,"Stress_rest2") %>% pivot_longer(.,names_to="time", names_prefix="Stress_",cols=starts_with("Stress_"), values_to="Stress")


plot_data <- plot_data%>%
  group_by(time, class_Stress) %>%
  summarize(n = n(),
            mean = mean(Stress,na.rm=T))


pd <- position_dodge(0.2)
ggplot(plot_data,
       aes(x = factor(plot_data$time, ordered=T,levels=c("rest1","iat", "conf","masc","rest2")),
                      y = mean, group=as.factor(class_Stress), color=as.factor(class_Stress))) +
  geom_point(position=pd,
             size=3) +
  geom_line(position=pd,
            linewidth = 1) +
    scale_color_brewer(palette="Set1") +
  theme_minimal() +
  labs(title = "Growth Curve trajectories of Stress",
       x = "",
       y = "",
       color = "Class")
```
