---
title: "Latent classes of HRV and their determinants"
author: "Setareh"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyr)
library(dplyr)
library(ggplot2)
library(ggpubr)
library(patchwork)
library(gmodels)
library(foreign)
library(gtsummary)
```

```{r data_entry}
dta_raw <- read.csv("raw40 subjects_HRV all conditions.csv", sep = ";")

# Non-physiological data
dta_nophysio <- read.spss(file="non physiol variables.sav", to.data.frame = TRUE, use.value.labels = TRUE)
dta_nophysio$subject <- as.numeric(as.character(dta_nophysio$subject))

# Class memberships from scripts 02, 03, 05 and 01 (PNS: single class, not used here)
add_class <- function(df, rdata, name) {
  env <- new.env()
  load(rdata, envir = env)
  cls <- env$pred[, c("subject", "class")]
  names(cls)[2] <- name
  merge(df, cls, by = "subject")
}

dta_raw <- dta_raw %>%
  add_class("lcmm_rMSSD.RData",  "class_rMSSD") %>%
  add_class("lcmm_SNS.RData",    "class_SNS") %>%
  add_class("lcmm_HR.RData",     "class_HR") %>%
  add_class("lcmm_Stress.RData", "class_Stress")

dta_raw <- dta_raw %>% rename(ESP_score= "ESPT_C_scores..threshold.49.", LSRP_total="LSRP_Total..psychopathy.in.non.institutionalised.populations..0.48..non.psychopathic..49.57.mixed....58.psychopathic.", AQ_total= "AQ_10_Total" ,SPQ_total= "SPQ_B_5_Total", RT_cI_iI="RT_cI.iI..if.positive..iI.is.shorter.meaning.efficient.adaptation.in.emotional.regulation." )

dta_nophysio <- merge(dta_nophysio,
                      dta_raw %>% dplyr::select(subject, class_rMSSD, class_SNS, class_HR, class_Stress,
                                                ESPT_total = ESP_score, cong_ACC, incong_ACC, RT_clil = RT_cI_iI),
                      by = "subject")
dta_nophysio <- dta_nophysio%>%rename(AQ_total= "AQ_10_Total" ,SPQ_total= "SPQ_B_5_Total")

# Class labels (class 1 of rMSSD and class 2 of SNS contain a single subject)
dta_nophysio$class_rMSSD[dta_nophysio$class_rMSSD==2] <-  "Low_rMSSD"
dta_nophysio$class_rMSSD[dta_nophysio$class_rMSSD==3] <-  "High_rMSSD"
dta_nophysio$class_SNS[dta_nophysio$class_SNS==1] <- "Low_SNS"
dta_nophysio$class_SNS[dta_nophysio$class_SNS==3] <- "High_SNS"

dta_nophysio$class_HR <- ifelse(dta_nophysio$class_HR==1, "Low_HR", "High_HR")

dta_nophysio$class_Stress<- ifelse(dta_nophysio$class_Stress==1, "Low_stress", "High_stress")

dta_IAT <- read.csv(file="IAT data.csv", sep = ";")%>%rename(subject=Subject)

dta_nophysio <- merge(dta_nophysio,dta_IAT,by="subject")

# Variables used in the descriptive tables
vars_socio <- c("age","BMI","socioeco","Illegal_act","laterality","stup_psych","alcool","drug_all","medicament")
vars_psycho <- c("ESPT_CSommeA", "ESPT_CSommeB","LSRP_Primary","LSRP_Secondary", "RPQ_React", "RPQ_Proact", "IRI_PT", "IRI_FS", "IRI_EC", "IRI_PD", "AQ_total", "SPQ_total")
vars_behav <- c("Final.IAT","cong_ACC","incong_ACC", "cong_RT", "incong_RT", "RT_clil", "MASC_TotCorrect","MASC_CogToMCorr", "MASC_AffToMCorr", "MASC_UnderToM", "MASC_OverToM")
vars_cont <- c("ESPT_CSommeA", "RPQ_Proact", "AQ_total", "SPQ_total")

# Table comparing two classes
class_table <- function(dat, by, vars, digits = 2) {
  args <- list(data = dat, by = by, type = list(all_of(vars_cont) ~ "continuous"),
               missing = "no", statistic = list(all_continuous() ~ "{mean} ({sd})"),
               include = vars)
  if (!is.null(digits)) args$digits <- list(all_continuous() ~ c(digits))
  do.call(tbl_summary, args) %>%
    add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2))
}

# -log10(p) for the Manhattan plot
p_values <- function(tbl, class_name) {
  data.frame(covariates = tbl[["table_body"]]$variable,
             p_value = -log10(tbl[["table_body"]]$p.value),
             class = class_name)
}

format_table <- function(tbl, header) {
  tbl %>%
    modify_header(label ~ "**Non Physiological Variables**") %>%
    modify_spanning_header(c("stat_1", "stat_2") ~ header) %>%
    modify_footnote(all_stat_cols() ~ "Mean (SD) or Frequency (%)") %>%
    bold_labels() %>% bold_p()
}
```


\newpage

# Descriptive statistics

## Overall sample

```{r}
dta_nophysio %>% tbl_summary(type = list(all_of(vars_cont) ~ "continuous"), missing = "no",
                             digits = list(all_continuous() ~ c(2)),
                             statistic = list(all_continuous() ~ "{mean} ({sd})"),
                             include = all_of(c(vars_socio, vars_psycho, vars_behav))) %>%
  modify_header(label ~ "**Non Physiological Variables Sample descriptive**") %>%
  modify_footnote(all_stat_cols() ~ "Mean (SD) or Frequency (%)") %>%
  bold_labels()
```


\newpage

## rMSSD classes

```{r}
dta_rmssd <- dta_nophysio %>% filter(class_rMSSD != 1)

class_table(dta_rmssd, "class_rMSSD", c(vars_socio, vars_psycho, vars_behav)) %>%
  format_table("**rMSSD_Classes**")

vis_desc_data <- p_values(class_table(dta_rmssd, "class_rMSSD", c(vars_psycho, vars_behav)), "rMSSD")
```


\newpage

## SNS classes

```{r}
dta_sns <- dta_nophysio %>% filter(class_SNS != 2)

class_table(dta_sns, "class_SNS", c(vars_socio, vars_psycho, vars_behav)) %>%
  format_table("**SNS_Classes**")

vis_desc_data <- rbind(vis_desc_data,
                       p_values(class_table(dta_sns, "class_SNS", c(vars_psycho, vars_behav)), "SNS"))
```


\newpage

## HR classes

```{r}
class_table(dta_nophysio, "class_HR", c(vars_socio, vars_psycho, vars_behav), digits = NULL) %>%
  format_table("**HR_Classes**")

vis_desc_data <- rbind(vis_desc_data,
                       p_values(class_table(dta_nophysio, "class_HR", c(vars_psycho, vars_behav), digits = NULL), "HR"))
```


\newpage

## Stress classes

```{r}
class_table(dta_nophysio, "class_Stress", c(vars_socio, vars_psycho, vars_behav)) %>%
  format_table("**Stress_Classes**")

vis_desc_data <- rbind(vis_desc_data,
                       p_values(class_table(dta_nophysio, "class_Stress", c(vars_psycho, vars_behav)), "Stress"))
```

\newpage

# Visualization

```{r common}
phases <- c("rest1","iat", "conf","masc","rest2")

# Mean and 95% CI per phase and class
summarise_ci <- function(dat, value, group) {
  dat %>%
    group_by(time, .data[[group]]) %>%
    summarize(n = n(),
              mean = mean(.data[[value]], na.rm = TRUE),
              sd = sd(.data[[value]], na.rm = TRUE),
              ci = qt(0.975, df = n - 1) * sd / sqrt(n),
              .groups = "drop") %>%
    mutate(time = factor(time, ordered = TRUE, levels = phases))
}

to_long <- function(dat, var, class_col = NULL) {
  cols <- paste0(var, "_", phases)
  dat %>% dplyr::select(all_of(c("subject", class_col, cols))) %>%
    pivot_longer(cols = all_of(cols), names_to = "time",
                 names_prefix = paste0(var, "_"), values_to = var)
}

mytheme <- theme(axis.text.y =  element_text(face="bold",size = 12),
                 axis.text.x = element_text(face="bold",size = 12),
                 legend.text = element_text(face="bold",size = 12),
                 legend.title = element_text(face="bold",size = 12))

pd <- position_dodge(0.2)
```


## Growth curve trajectories with one-member classes

```{r rmssd_outlier}
plot_data_rmssd <- to_long(dta_raw, "rMSSD", "class_rMSSD")

plot_data_rmssd$class_rMSSD[plot_data_rmssd$class_rMSSD==1] <- "One Subject"
plot_data_rmssd$class_rMSSD[plot_data_rmssd$class_rMSSD==2] <- "Low_rMSSD"
plot_data_rmssd$class_rMSSD[plot_data_rmssd$class_rMSSD==3] <- "High_rMSSD"
plot_data_rmssd$class_rMSSD <-factor(plot_data_rmssd$class_rMSSD, levels=c("Low_rMSSD","High_rMSSD","One Subject"))
plot_data_rmssd <- summarise_ci(plot_data_rmssd, "rMSSD", "class_rMSSD")

p1_rmssd <- ggplot(plot_data_rmssd, aes(x = time, y = mean, group = class_rMSSD, color = class_rMSSD)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  scale_color_manual(values=c("#377EB8","#4DAF4A", "red"))  +
  theme_minimal() +
  labs(x = "Phases", y = "rMSSD(ms)", color = "Latent Class") +
  geom_hline(yintercept = c(27,57), linetype = "dashed") +
  annotate("text", y = 60, x = 2, label = "General Norm", color = "black")
```


```{r SNS_outlier}
plot_data_SNS <- to_long(dta_raw, "SNS", "class_SNS")

plot_data_SNS$class_SNS[plot_data_SNS$class_SNS==1] <- "Low"
plot_data_SNS$class_SNS[plot_data_SNS$class_SNS==2] <- "One Subject"
plot_data_SNS$class_SNS[plot_data_SNS$class_SNS==3] <- "High"
plot_data_SNS <- summarise_ci(plot_data_SNS, "SNS", "class_SNS")

p1_SNS <- ggplot(plot_data_SNS, aes(x = time, y = mean, group = class_SNS, color = class_SNS)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  scale_color_manual(values=c("#FF7F00","#A65628","red"))  +
  theme_minimal() +
  labs(x = "Phases", y = "SNS activity", color = "Latent Class") +
  geom_hline(yintercept = c(-1,1), linetype = "dashed") +
  annotate("text", y = 1.2, x = 2, label = "General Norm", color = "black")
```

```{r ensemble_outlier,fig.width=10,fig.height=5}
graph <- ggarrange(p1_rmssd + mytheme, p1_SNS + mytheme, labels=c("(A) rMSSD", "(B) SNS"),align="hv", nrow = 1,ncol=2,font.label = list(size = 14, color = "black", face = "bold"))

graph <- annotate_figure(graph, top = text_grob("Growth Curve trajectories", face = "bold", size = 14))
graph

pdf(file = "supp_figure1.pdf", width = 10, height = 5)
print(graph)
invisible(dev.off())
```

## Growth curve trajectories with 95% confidence intervals


```{r rmssd_without}
plot_data_rmssd <- to_long(dta_raw, "rMSSD", "class_rMSSD") %>% filter(class_rMSSD != 1)

plot_data_rmssd$class_rMSSD <- ifelse(plot_data_rmssd$class_rMSSD == 2, "Low_rMSSD", "High_rMSSD")
plot_data_rmssd$class_rMSSD <- factor(plot_data_rmssd$class_rMSSD, levels=c("Low_rMSSD","High_rMSSD"))
plot_data_rmssd <- summarise_ci(plot_data_rmssd, "rMSSD", "class_rMSSD")

p2_rmssd <- ggplot(plot_data_rmssd, aes(x = time, y = mean, group = class_rMSSD, color = class_rMSSD)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = .1, position=pd, linewidth=1) +
  scale_color_manual(values=c("#377EB8","#4DAF4A"))  +
  theme_minimal() +
  labs(x = "Phases", y = "rMSSD(ms)", color = "Latent Classes") +
  geom_hline(yintercept = c(27,57), linetype = "dashed") +
  annotate("text", y = 58, x = 2, label = "General Norm", color = "black")
```


```{r sns_without}
plot_data_SNS <- to_long(dta_raw, "SNS", "class_SNS") %>% filter(class_SNS != 2)

plot_data_SNS$class_SNS <- ifelse(plot_data_SNS$class_SNS == 1, "Low_SNS", "High_SNS")
plot_data_SNS <- summarise_ci(plot_data_SNS, "SNS", "class_SNS")

p2_SNS <- ggplot(plot_data_SNS, aes(x = time, y = mean, group = class_SNS, color = class_SNS)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = .1, position=pd, linewidth=1) +
  scale_color_manual(values=c("#FF7F00","#A65628"))  +
  theme_minimal() +
  labs(x = "Phases", y = "SNS activity", color = "Latent Classes") +
  geom_hline(yintercept = c(-1,1), linetype = "dashed") +
  annotate("text", y = 1.2, x = 2, label = "General Norm", color = "black")
```


```{r hr_without}
plot_data_HR <- to_long(dta_raw, "HR", "class_HR")

plot_data_HR$class_HR <- ifelse(plot_data_HR$class_HR==1, "Low_HR", "High_HR")
plot_data_HR <- summarise_ci(plot_data_HR, "HR", "class_HR")

p2_HR <- ggplot(plot_data_HR, aes(x = time, y = mean, group = class_HR, color = class_HR)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = .1, position=pd, linewidth=1) +
  scale_color_manual(values=c("#E41A1C","#FFFF33"))  +
  theme_minimal() +
  labs(x = "Phases", y = "Heart Rate(bpm)", color = "Latent Classes") +
  geom_hline(yintercept = c(63,95), linetype = "dashed") +
  annotate("text", y = 96, x = 2, label = "General Norm", color = "black")
```

```{r stress_without}
plot_data_stress <- to_long(dta_raw, "Stress", "class_Stress")

plot_data_stress$class_Stress <- ifelse(plot_data_stress$class_Stress==1, "Low_stress", "High_stress")
plot_data_stress$class_Stress <- relevel(as.factor(plot_data_stress$class_Stress), ref="Low_stress")
plot_data_stress <- summarise_ci(plot_data_stress, "Stress", "class_Stress")

p2_Stress <- ggplot(plot_data_stress, aes(x = time, y = mean, group = class_Stress, color = class_Stress)) +
  geom_point(position=pd, size=3) +
  geom_line(position=pd, linewidth = 1) +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = .1, position=pd, linewidth=1) +
  scale_color_manual(values=c("#F781BF","#984EA3"))  +
  theme_minimal() +
  labs(x = "Phases", y = "Stress", color = "Latent Classes") +
  geom_hline(yintercept = c(10,15), linetype = "dashed") +
  annotate("text", y = 15.5, x = 4.5, label = "General Norm", color = "black")
```


```{r pns}
# PNS: single latent class, whole sample
plot_data_PNS <- to_long(dta_raw, "PNS") %>% mutate(all = "All")
plot_data_PNS <- summarise_ci(plot_data_PNS, "PNS", "all")

p2_PNS <- ggplot(plot_data_PNS, aes(x = time, y = mean, group = 1)) +
  geom_point(size=3, color="#999999") +
  geom_line(linewidth=1, color="#999999") +
  geom_errorbar(aes(ymin = mean - ci, ymax = mean + ci), width = .1, linewidth=1, color="#999999") +
  theme_minimal() +
  labs(x = "Phases", y = "PNS activity") +
  geom_hline(yintercept = c(-1,1), linetype = "dashed") +
  annotate("text", y = 1.1, x = 3, label = "General Norm", color = "black")
```


```{r ensemble, fig.height=15,fig.width=10}
plotfinal <- ((p2_HR + mytheme) + (p2_rmssd + mytheme)) /
             ((p2_SNS + mytheme) + (p2_PNS + mytheme)) /
             ((p2_Stress + mytheme) + plot_spacer())

plotfinal + plot_annotation(title = "Growth Curve trajectories",
                            subtitle = "Vertical bars denote 95% Confidence Intervals",
                            tag_levels = list(c("(A) HR","(D) rMSSD","(B) SNS","(E) PNS","(C) Stress"))) &
  theme(plot.title = element_text(hjust = 0.5, vjust = 2),
        plot.subtitle = element_text(hjust = 0.5, vjust = 2),
        plot.tag.position = c(0, 1),
        plot.tag = element_text(size = 10, face="bold", hjust = 0.1, vjust = -0.5))
```

\newpage

## Manhattan plot of psychometric and behavioral variables

```{r manhattanplot}
vis_desc_data$covariates <- factor(vis_desc_data$covariates, levels = c(vars_psycho, vars_behav))

vis_data_psychometric <- vis_desc_data %>% filter(covariates %in% vars_psycho)
vis_data_behavioral <- vis_desc_data %>% filter(covariates %in% vars_behav)

p1 <- ggplot(vis_data_psychometric,aes(x=covariates, y=p_value,color=class))+ geom_point()+geom_hline(yintercept=-log10(0.05), lty="dashed")+geom_hline(yintercept=-log10(0.1), lty="dashed",color="grey")+labs(x="",y="-Log(P_value)") +theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position="bottom")+ggtitle("Psychometric measures Discriminant")+scale_color_discrete("Between latent classes of:")+scale_x_discrete(labels=c("Trauma", "PTSD suspicion", "Primary psychopathy", "Secondary psychopathy", "Reactive aggression", "Proactive aggression","Perspective taking","Fantasy scale","Empathic concern","Personal distress","Autistic traits","Schizotypy"))+ylim(c(0,3))

p2 <- ggplot(vis_data_behavioral,aes(x=covariates, y=p_value,color=class))+ geom_point()+geom_hline(yintercept=-log10(0.05), lty="dashed")+geom_hline(yintercept=-log10(0.1), lty="dashed",color="grey")+labs(x="",y="-Log(P_value)") +theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),legend.position="bottom")+ggtitle("Behavioral measures Discriminant")+scale_color_discrete("Between latent classes of:")+scale_x_discrete(labels=c("Final IAT", "Accuracy (congruent)", "Accuracy (incongruent)","RT (congruent)","RT (incongruent)","Conflict adaptation","Total ToM","Cognitive ToM","Affective ToM","Under ToM","Over ToM"))+ylim(c(0,3))

ggarrange(p1,p2,common.legend = TRUE, legend="bottom")

ggplot(vis_desc_data,aes(x=covariates, y=p_value,color=class))+ geom_point()+geom_hline(yintercept=-log10(0.05), lty="dashed")+labs(x="Psychometric and Behavioral measures",y="-Log(P_value)") +theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+ggtitle("Discriminant characteristics of the Latent Classes")+scale_color_discrete("Between latent classes of:")
```


\newpage

# Cross-Tabulation and Chi-square test

## between rMSSD and SNS classes

```{r}
with(dta_raw, CrossTable(class_rMSSD,class_SNS, digits = 2, chisq = T,simulate.p.value=F ))
```

\newpage

## between HR and SNS classes

```{r}
with(dta_raw, CrossTable(class_HR,class_SNS, digits = 2, chisq = T,simulate.p.value=F ))
```


\newpage

## between rMSSD and Stress classes

```{r}
with(dta_raw, CrossTable(class_rMSSD,class_Stress, digits = 2, chisq = T,simulate.p.value=F))
```


\newpage

## between HR and Stress classes

```{r}
with(dta_raw, CrossTable(class_HR,class_Stress, digits = 2, chisq = T,simulate.p.value=F))
```


\newpage

## between SNS and Stress classes

```{r}
with(dta_raw, CrossTable(class_SNS,class_Stress, digits = 2, chisq = T,simulate.p.value=F))
```

\newpage

## between rMSSD and HR classes

```{r}
with(dta_raw, CrossTable(class_rMSSD,class_HR, digits = 2, chisq = T,simulate.p.value=F))
```

\newpage

# Session information

```{r session_info, echo=TRUE}
sessionInfo()
```
